Bayesian Statistical Modeling with Predictors from LLMs

### 2024-05-21 04-paper/paper-LLM-probmod.tex
integrated feedback / typos

\ifuniqueAffiliation % Standard variant of author block
  \author{ Michael Franke\thanks{Corresponding author.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{mchfranke@uni-tuebingen.de} \\
	%% examples of more authors
	\And
	Polina Tsvilodub \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{polina.tsvilodub@gmail.com} \\
	\And
	Fausto Carcassi \\
	ILLC\\
	University of Amsterdam\\
	\texttt{fausto.carcassi@gmail.com} \\
}
\else
% Multiple affiliations variant of author block
\usepackage{authblk}
\renewcommand\Authfont{\bfseries}
\setlength{\affilsep}{0em}
% box is needed for correct spacing with authblk
% \newbox{\orcid}\sbox{\orcid}{\includegraphics[scale=0.06]{orcid.pdf}}
\author{Michael Franke, Polina Tsvilodub, Fausto Carcassi}
\affil{Department of Linguistics\\University of Tübingen\\
\texttt{[michael.franke|polina.tsvilodub|fausto.carcassi]@uni-tuebingen.de}}
\fi

@@

\begin{abstract}
  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
  This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of Bayesian statistical modeling.
  % We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
  % Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
\end{abstract}

@@

Data from human participants for experiments of this kind usually show some variability between items, and also variability between participants.
This variability is commonly incorporated in standard statistical models as random stochastic variation, e.g., by using hierarchical regression models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}, as shown on the left side of Figure~\ref{fig:stats-model-types}.
Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the model's predictions will naturally be set-up to predict data by taking condition-level properties, possibly in conjunction with item-level properties, into account \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, as shown schematically in the middle of Figure~\ref{fig:stats-model-types}.
On the other hand, LLMs first and foremost provide predictions about each item.
While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested, thus raising concerns about robustness of predictions under perturbations of input prompts \mf{cite}.

These considerations raise two important conceptual and empirical questions.
First, we need to ask whether the item-level predictions made by LLMs are empirically correct.
Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
To investigate these issues, this paper introduces different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.,][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.
For reference games, a widely used probabilistic cognitive model exists, namely the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.

@@

  \caption{Structure of a reference game with human participants. Each trial consists of a set of objects, the so-called context. In production trials, participants choose a single word to describe a trigger object from the context. In interpretation trials, an object is selected as the likely object a trigger word referring to.}
  \label{fig:ref-game}
\end{figure}

@@

The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where human choices are informative about the pragmatic reasoning that human decision makers engage in.
In this example, there are two features that differ across three objects (here shape and color).
One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
%
In a critical production trial, the trigger object to describe is one of the two objects with a unique feature.
The speaker has four words to choose from.
The \textbf{target utterance} is the word which uniquely describes the trigger object.
The \textbf{distractor utterance} is the word that is true of the trigger object, but also true of another object.
The other utterances, both of which are false of the trigger object are \textbf{competitor utterance}.
%
In a critical interpretation trial, the trigger word is one that is true of two of the three objects.
If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
The \textbf{competitor object} is the other object of which the trigger word is true.
The \textbf{distractor object} is the object of which the trigger word is false.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model.
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

\paragraph{The Rational Speech Act model.}
As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{\text{prior}}(s)$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u)\  P_{\text{prior}}(s)\,.
\end{align*}
%
The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object:
%
\begin{align}
  \label{eq:RSA-speaker}
  P_{S}(u \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u) \right ] \,.
\end{align}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align}
  \label{eq:RSA-interpreter}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P_{\text{prior}}(s) \,.
\end{align}

@@

\paragraph{Condition-level predictions.}
In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation).

@@

Based on each option's score, we can define the \emph{item-level prediction} of choosing option $y_{ki}$ in term of soft-maximization as:
%
\begin{align*}
P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{kl} \right)\,.
\end{align*}
%
Notice that the usual item-level prediction used in benchmark testing is actually a ``winner-takes-all'' strategy, but this is just a special case of the above for $\alpha_{C} \rightarrow \infty$.
Here, the $\alpha_{C}$ parameter corresponds to inverse temperature, so that $\alpha_{C} \rightarrow \infty$ corresponds to greedy sampling with temperature zero.
Like in the RSA model, we use an independent soft-max parameter for each condition.

@@

To address this question, we assessed item-level scores $\text{S}_{kl}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{kl} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{00-pics/GPT-closeness-target-by-alpha-item-level.pdf}
  \caption{
    Predicted probability of highest-scoring answer category, averaged over items, for different values of softmax parameter $\alpha$ (black lines).
    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
  }
  \label{fig:closeness-target-item-level}
\end{figure}

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM-based model predicts variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.

@@

The \emph{average-scores} and \emph{average-probabilities} predictors are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
However, based on the results from the previous section, the item-level predictions for the WTA strategy are demonstrably incongruent with the human data.
In general, the average-probability and average-WTA predictors can lead to qualitatively different results about task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).

@@

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} model are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
  \label{fig:LLaMA-results-vPPC}
\end{figure}

>>>>

\ifuniqueAffiliation % Standard variant of author block
  \author{ Michael Franke\thanks{Corresponding author.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{mchfranke@gmail.com} \\
	%% examples of more authors
	\And
	Polina Tsvilodub \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{polina.tsvilodub@gmail.com} \\
	\And
	Fausto Carcassi \\
	ILLC\\
	University of Amsterdam\\
	\texttt{fausto.carcassi@gmail.com} \\
}
\else
% Multiple affiliations variant of author block
\usepackage{authblk}
\renewcommand\Authfont{\bfseries}
\setlength{\affilsep}{0em}
% box is needed for correct spacing with authblk
% \newbox{\orcid}\sbox{\orcid}{\includegraphics[scale=0.06]{orcid.pdf}}
\author{Michael Franke, Polina Tsvilodub, Fausto Carcassi}
\affil{Department of Linguistics\\University of Tübingen\\
\texttt{[michael.franke|polina.tsvilodub|fausto.carcassi]@uni-tuebingen.de}}
\fi

@@

\begin{abstract}
  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
  This raises questions about the human-likeness of LLM-derived information, alignment with human intuition in larger applications, and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of Bayesian statistical modeling.
  % We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
  % Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs do not capture the variance in the human data at the item-level.
  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to human data.
  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at the aggregate condition-level, for which they are, however, not designed to make predictions in the first place.
\end{abstract}

@@

Data from human participants for experiments of this kind usually show some variability between items, and also variability between participants.
This variability is commonly incorporated in standard statistical models as random stochastic variation, e.g., by using hierarchical regression models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}, as shown on the left side of Figure~\ref{fig:stats-model-types}.
Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the model's predictions will naturally be set-up to predict data by taking condition-level properties, possibly in conjunction with item-level properties, into account \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, as shown schematically in the middle of Figure~\ref{fig:stats-model-types}.
On the other hand, LLMs first and foremost provide predictions about each item.
While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested.
Consequently, this raises concerns about robustness of predictions under perturbations of input prompts \mf{cite}.

These considerations raise two important conceptual and empirical questions.
First, we need to ask whether the item-level predictions made by LLMs are empirically correct, i.e., match the human data at the item-level.
Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
To investigate these issues, this paper introduces different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.,][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.
For reference games, a widely used probabilistic cognitive model exists, namely the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.

@@

  \caption{Structure of a reference game with human participants. Each trial consists of a set of objects, the so-called context. In production trials, participants choose a single word to describe a trigger object from the context. In interpretation trials, an object is selected as the likely object a trigger word is referring to.}
  \label{fig:ref-game}
\end{figure}

@@

The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where human choices are informative about the pragmatic reasoning that human decision makers engage in.
In this example, there are two features that differ across three objects (here shape and color).
One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
%
In a critical production trial, the trigger object to describe is one of the two objects with a unique feature.
The speaker has four words to choose from.
The \textbf{target utterance} is the word which uniquely describes the trigger object.
The \textbf{competitor utterance} is the word that is true of the trigger object, but also true of another object.
The other utterances, both of which are false of the trigger object are \textbf{distractor utterance}.
%
In a critical interpretation trial, the trigger word is the one that is true of two of the three objects.
If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
The \textbf{competitor object} is the other object of which the trigger word is true.
The \textbf{distractor object} is the object of which the trigger word is false.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model.
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over states (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a parameterized stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

\paragraph{The Rational Speech Act model.}
As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies and Bayesian inference, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{\text{prior}}(s)$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u)\  P_{\text{prior}}(s)\,.
\end{align*}
%
The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object given the utterance:
%
\begin{align}
  \label{eq:RSA-speaker}
  P_{S}(u \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u) \right ] \,.
\end{align}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align}
  \label{eq:RSA-interpreter}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P_{\text{prior}}(s) \,.
\end{align}

@@

\paragraph{Condition-level predictions.}
In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation) for a given $\alpha_{c}$.

@@

Based on each option's score, we can define the \emph{item-level prediction} of choosing option $y_{kl}$ in terms of soft-maximization as:
%
\begin{align*}
P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{kl} \right)\,.
\end{align*}
%
Notice that the usual item-level prediction used in benchmark testing is actually a ``winner-takes-all'' strategy, which would choose any option that maximizes the score, but this is just a special case of the above for $\alpha_{C} \rightarrow \infty$.
Here, the $\alpha_{C}$ parameter corresponds to inverse temperature, so that $\alpha_{C} \rightarrow \infty$ corresponds to greedy sampling with temperature zero.
Like in the RSA model, we use an independent soft-max parameter for each condition.

@@

To address this question, we assessed item-level scores $\text{S}_{kl}$ from the text-davinci-003 instance of GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we treat the union of the distractor words as a single option.}
Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = (1- \epsilon_{c}) \  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
With priors on parameters as for the RSA model described in Section~\ref{sec:model-pred-from}, the resulting likelihood function for item-level data with item-level predictions is:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{kl} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{00-pics/GPT-closeness-target-by-alpha-item-level.pdf}
  \caption{
    Predicted probability of highest-scoring answer category, averaged over items, for different values of softmax parameter $\alpha$ (black lines).
    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
    For reference, the target probability under a random choice strategy and under the ``winner-takes-all'' (WTA) strategy are shown with dashed lines.
    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
  }
  \label{fig:closeness-target-item-level}
\end{figure}

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM-based model predicts variance which is not borne out by the human data.
Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that uses the most obvious item-level scores, derived from the predicted log-probabilities, to predict what human participants choose on a by-item level is ruled out by the current experimental data.

@@

The \emph{average-scores} and \emph{average-probabilities} predictors are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
However, based on the results from the previous section, the item-level predictions for the WTA strategy are demonstrably incongruent with the human data.
In general, the average-probability and average-WTA predictors can lead to qualitatively different results for task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).

@@

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} models are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies. \mf{update plot by replacing 'hf' with 'base'.}}
  \label{fig:LLaMA-results-vPPC}
\end{figure}

All previous results where obtained for LMM predictions derived from GPT-3.5 davinci.
To investigate whether key results replicate for other LLMs and scales, we ran the same analyses also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter base models (marked as `base'), and the models of similar size fine-tuned on chat data (marked as `chat').
Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
The plots show the \emph{differences} between observed counts and the models' posterior predictions for expected counts for each condition and response option.
We find that the average-WTA predictor is consistently able to capture the data from the production condition, except for one of the six models (LLaMA2-chat 70B).
When the average-WTA predictor passes the model checking criterion for the production data, the posteriors for the power-law parameter $\alpha$ are credibly different from 1, thus suggesting that, contrary to the results for GPT-3.5 a mere ``WTA average'' is not a good prediction strategy for the production data for the LLaMA2-based models.
Looking at interpretation, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 model.
Finally, for item-level data, as for GPT-3.5, no LLaMA2-based model captured the item-level variance observed in the human data.
In sum, we find general support for the previous conclusions that the average-WTA predictor is most successful in capturing the condition-level data, and therefore that, where models manage to recover the condition-level data, they seem to be ``right for the wrong reasons.''
Additionally, we also find variability in which method of aggregation works well with which LLM as scoring model.
